### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
removing spainv-ilut section

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we choose
\textit{block Jacobi} preconditioning as a left preconditioner for the
system. Like point Jacobi preconditioning, block Jacobi
preconditioning extracts the diagonal elements of the matrix as the
preconditioner where now the elements extracted are the blocks on the
diagonal as shown on the left side of
Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
trivial with each diagonal block inverted separately. For the $SP_N$
equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
$N_g\times(N+1)/2$.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks of the matrix. Right: Inversion of the
      preconditioner is trivial and decoupled by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}
In addition, in the limit of a block size of one, the block Jacobi
method reduces to the point Jacobi method.

@@

Spectral radius computations for the block Jacobi preconditioned
iteration matrix were performed for the same problems as the point
Jacobi preconditioned case. Table~\ref{tab:group1bj} gives the results
for the 1-group case, Table~\ref{tab:group10dsbj} for the 10-group
case with full downscatter only and Table~\ref{tab:group10usbj} for
the 10-group case with full downscatter and full upscatter. A block
size of $N_g\times(N+1)/2$ was used for the preconditioner in all
cases.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1bj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dsbj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10usbj}
\end{table}

@@

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
computation\footnote{The operator will change if, for example, physics
  feedback through temperature or potentially burnup is
  considered. These additional physics will modify the cross sections
  used to assemble $\mathbf{M}$. The calculations presented here will
  consist of a single eigenvalue calculation with a static
  $\mathbf{M}$.} while the addition of fission to the system will only
modify the source of neutrons and the multiplication factor at every
eigenvalue iteration while not affecting Monte Carlo transport.

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
Based on the success of block Jacobi preconditioning with the test
problem used for the spectral radius parameter study, we use it first
to solve the $17 \times 17$ fuel assembly problem. A single energy
group problem was first solved with $SP_1$ discretization, effectively
giving the one-speed neutron diffusion system for the fuel assembly
resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. Even if
the problem may be ill-conditioned, we do expect convergence of MCSA
given that the spectral radius requirement was met. To investigate
further, a block Jacobi preconditioned Richardson iteration was used
to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor convergence is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

@@

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
For the fuel assembly problem, the initial block Jacobi
preconditioned\footnote{For 1-group $SP_1$ problems the block size is
  1, giving a point Jacobi preconditioner.}  calculations used a
one-speed $SP_1$ discretization, effectively giving a diffusion
system. To study the breakdown of MCSA at iteration matrix spectral
radii near one, we will use the simpler homogeneous 2-dimensional
one-speed neutron diffusion system presented in
Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
this system, we can vary the cross sections while maintaining a fixed
grid in order to achieve varying spectral radii. For these studies, we
neglect fission as MCSA behavior is dictated by the transport operator
$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
generate a fixed source.

@@

%%---------------------------------------------------------------------------%%
\section{Advanced Preconditioning Strategies}
\label{subsec:spn_advanced_preconditioning}
For the fuel assembly criticality problem presented in the previous
section, the spectral radius was near one using Jacobi preconditioning
and in a regime in which MCSA breakdown is observed. In this regime
the number of stochastic histories required to converge MCSA increases
rapidly and the resulting poor performance is compounded by those
histories being increasingly expensive to compute. To overcome this
difficulty, advanced preconditioning strategies for the $SP_N$ problem
are required beyond simple Jacobi methods that can reduce the spectral
radius into a region of better MCSA performance. Several modern
algebraic preconditioning strategies will be presented here and used
within the explicit preconditioning framework given by
Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
solutions of the fuel assembly criticality problem will be presented.

\subsection{ILUT Preconditioning}
\label{subsec:spn_ilut_preconditioning}
Incomplete lower-upper (ILU) factorizations of the linear operator can
be used a simple mechanism to form an approximate inverse of a
preconditioner. To build the factorization, the sparse upper and lower
triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
that the residual matrix formed by the factorization
\cite{saad_iterative_2003}:
\begin{equation}
  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
  \label{eq:ilu_residual_matrix}
\end{equation}
has a specified sparsity pattern and element magnitude. When a
magnitude threshold is used, elements generated in the factors below
that magnitude are dropped, resulting in ILU threshold (ILUT)
preconditioning. The sparsity pattern in this case is determined from
the input matrix to be preconditioned and the number of elements
maintained in the factor is specified by a fill level parameter. A
fill level of 1 will generate the same number of elements as the
sparsity pattern of the input matrix while a fill level of 2 will
contain twice as many elements in the factorization, resulting in a
better representation of the true LU factorization. The inverse of the
lower and upper triangular factors may then be easily inverted by
means of simple elimination to produce the preconditioner. For MCSA,
$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
the right to precondition the system.

@@

\subsection{Sparse Approximate Inverse Preconditioning}
\label{subsec:spn_spainv_preconditioning}
Explicit formation of the preconditioner inverse is a requirement of
the current MCSA preconditioning strategy. As a result, although ILUT
preconditioning provided excellent iterative performance, a
significant time penalty is paid to extract the inverse of the
triangular factors through $N$ triangular solves. In addition, this
explicit inversion yielded a composite linear operator that was dense,
requiring modification of the system in order to achieve better
scalability and Monte Carlo performance. Sparse approximate inverse
(SPAINV) preconditioning is a technique that may potentially alleviate
both of these constraints at the cost of reduced iterative
performance. The method produces a sparse approximation of the inverse
matrix directly by minimizing the Frobenius norm of the residual
matrix \cite{saad_iterative_2003}:
\begin{equation}
  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
  \label{eq:frobenius_norm_min}
\end{equation}
where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
$j^{th}$ columns of the identity and preconditioning matrices
respectively. Here, $\mathbf{M}$ is the actual inverse of the
preconditioner and is formed directly by the algorithm. On the right
hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
represented as a effective sum of $N$ linear system residual norms.

@@

\subsection{SPAINV Improved ILUT Preconditioning}
\label{subsec:spn_spainv_preconditioning}
As a means to improve convergence over using only SPAINV
preconditioning, ILUT preconditioning may be modified through an
additional application of SPAINV
(SPAINV-ILUT)\cite{saad_iterative_2003}. For the ILUT preconditioning
case we have the following linear problem to solve for the substituted
variable:
\begin{equation}
  \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
  \mathbf{L}^{-1} \mathbf{b}\:.
  \label{eq:ilut_standalone}
\end{equation}
We then apply SPAINV and minimize the following problem, this time on
the left:
\begin{equation}
  || \mathbf{I} - \mathbf{M} \mathbf{L}^{-1} \mathbf{A}
  \mathbf{U}^{-1} ||^2_F\:,
  \label{eq:ilut_minimize}
\end{equation}
giving a new preconditioned system:
\begin{equation}
  \mathbf{M} \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
  \mathbf{M} \mathbf{L}^{-1} \mathbf{b}\:.
  \label{eq:ilut_spainv}
\end{equation}
Doing this permits a smaller fill level to be used with ILUT and a
lower-order sparsity pattern to be used with SPAINV, giving marginally
more sparsity in the resulting composite linear operator while
maintaining good iterative performance.

MCSA preconditioned with SPAINV-ILUT was also used to solve the same
fuel assembly problem as with the other preconditioners for the same
1-group $SP_1$ discretization. To study convergence sensitivity to
SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
were parametrically varied with the number of iterations to converge a
single eigenvalue iteration for the fuel assembly problem recorded
along with the maximum number of non-zero entries observed in all
matrix rows for the left/right preconditioned composite linear
operator. SPAINV parameters for the data presented here were fixed as
it was found that the preconditioning quality was largely insensitive
to their variation. In addition, the fact that SPAINV is being used to
preconditioned a problem already preconditioned with ILUT means that
the input matrix sparsity pattern is already somewhat dense and large
SPAINV levels would result in an even denser system. Therefore, a
smaller SPAINV sparsity pattern level will be used for this
preconditioning. For each calculation, the number of iterations
required to converge reported was for a single eigenvalue iteration.
Again, \sn{3}{4} histories are used at each MCSA iteration to compute
the Monte Carlo correction using the adjoint collision estimator. All
SPAINV-ILUT calculations reported here were performed on a single CPU.

Figure~\ref{fig:spainv_ilut_iterations} gives the number of iterations
required to converge as a function of ILUT drop tolerance for ILUT
fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
threshold of 1.0 and sparsity level of 1. SPAINV threshold values
higher than 1.0 resulted in a loss of convergence and lower values did
not significantly affect performance. Larger sparsity levels resulted
in a composite operator that was too dense and a prohibitive amount of
CPU time required to compute the preconditioner. At smaller levels of
fill, the benefits of using an additional step of SPAINV
preconditioning are readily observed with the number of iterations
required to converge cut in half for large drop tolerances. For
smaller drop tolerances and higher ILUT levels of fill, the additional
step of SPAINV preconditioning offers marginal improvement to
iterative performance. As shown in Figure~\ref{fig:spainv_ilut_size},
because ILUT is used in the preconditioning sequence, sparsity is
again lost with only marginally improved non-zero entry numbers due to
the application of the SPAINV threshold. As a result, the quality
metrics given by Figure~\ref{fig:spainv_ilut_quality} are only
marginally better than those observed for ILUT preconditioning and
much larger than the SPAINV quality metric values.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge a
      single eigenvalue iteration for the fuel assembly problem with
      SPAINV-ILUT preconditioning as a function of ILUT drop
      tolerance.} \textit{Each colored curve represents the row size
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
      4.0, and 5.0 were used.}}
  \label{fig:spainv_ilut_iterations}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_size.pdf}
  \end{center}
  \caption{\textbf{Maximum number of non-zero entries observed for all
      rows in the composite linear operator for the fuel assembly
      problem with SPAINV-ILUT preconditioning given as a function of
      ILUT drop tolerance.} \textit{Each colored curve represents the
      row size for a different ILUT fill level. Fill levels of 1.5,
      2.0, 3.0, 4.0 and 5.0 were used.}}
 \label{fig:spainv_ilut_size}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_quality.pdf}
  \end{center}
  \caption{\textbf{SPAINV-ILUT preconditioning quality metric for the
      fuel assembly problem given as a function of ILUT drop
      tolerance.} \textit{Each colored curve represents the row size
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
      4.0 and 5.0 were used.}}
  \label{fig:spainv_ilut_quality}
\end{figure}

>>>>

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we choose
\textit{block Jacobi} preconditioning as a left preconditioner for the
system. Block Jacobi preconditioning extracts the diagonal elements of
the matrix as the preconditioner where now the elements extracted are
the blocks on the diagonal as shown on the left side of
Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
trivial with each diagonal block inverted separately. For the $SP_N$
equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
$N_g\times(N+1)/2$.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks of the matrix. Right: Inversion of the
      preconditioner is trivial and decoupled by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}

@@

Spectral radius computations for the block Jacobi preconditioned
iteration matrix were performed with Table~\ref{tab:group1bj} giving
the results for the 1-group case, Table~\ref{tab:group10dsbj} for the
10-group case with full downscatter only and
Table~\ref{tab:group10usbj} for the 10-group case with full
downscatter and full upscatter. A block size of $N_g\times(N+1)/2$ was
used for the preconditioner in all cases.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1bj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dsbj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10usbj}
\end{table}

@@

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
eigenvalue computation while the addition of fission to the system
will only modify the source of neutrons and the multiplication factor
at every eigenvalue iteration while not affecting Monte Carlo
transport.

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
We first use block Jacobi preconditioning to solve the $17 \times 17$
fuel assembly problem. A single energy group problem was first
computed with $SP_1$ discretization, effectively giving the one-speed
neutron diffusion system for the fuel assembly resulting in 20,088
degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. Even if
the problem may be ill-conditioned, we do expect convergence of MCSA
given that the spectral radius requirement was met. To investigate
further, a block Jacobi preconditioned Richardson iteration was used
to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor convergence is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

@@

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
For the fuel assembly problem, the initial block Jacobi preconditioned
calculations used a one-speed $SP_1$ discretization, effectively
giving a diffusion system. To study the breakdown of MCSA at iteration
matrix spectral radii near one, we will use the simpler homogeneous
2-dimensional one-speed neutron diffusion system presented in
Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
this system, we can vary the cross sections while maintaining a fixed
grid in order to achieve varying spectral radii. For these studies, we
neglect fission as MCSA behavior is dictated by the transport operator
$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
generate a fixed source.

@@

%%---------------------------------------------------------------------------%%
\section{Advanced Preconditioning Strategies}
\label{subsec:spn_advanced_preconditioning}
For the fuel assembly criticality problem presented in the previous
section, the spectral radius was near one using Jacobi preconditioning
and in a regime in which MCSA breakdown is observed. In this regime
the number of stochastic histories required to converge MCSA increases
rapidly and the resulting poor performance is compounded by those
histories being increasingly expensive to compute. To overcome this
difficulty, advanced preconditioning strategies for the $SP_N$ problem
are required beyond simple Jacobi methods that can reduce the spectral
radius into a region of better MCSA performance. Two modern algebraic
preconditioning strategies, ILUT and SPAINV, will be presented here
and used within the explicit preconditioning framework given by
Eq~(\ref{eq:left_right_mcsa}). Data
showing their effects on MCSA solutions of the fuel assembly
criticality problem will be presented.

\subsection{ILUT Preconditioning}
\label{subsec:spn_ilut_preconditioning}
ILUT preconditioning is chosen due to its applicability to problems
that have a dominating elliptic compoenent such as the fuel assembly
problem and because it is easily formulated algebraicly. Incomplete
lower-upper (ILU) factorizations of the linear operator can be used a
simple mechanism to form an approximate inverse of a
preconditioner. To build the factorization, the sparse upper and lower
triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
that the residual matrix formed by the factorization
\cite{saad_iterative_2003}:
\begin{equation}
  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
  \label{eq:ilu_residual_matrix}
\end{equation}
has a specified sparsity pattern and element magnitude. When a
magnitude threshold is used, elements generated in the factors below
that magnitude are dropped, resulting in ILU threshold (ILUT)
preconditioning. The sparsity pattern in this case is determined from
the input matrix to be preconditioned and the number of elements
maintained in the factor is specified by a fill level parameter. A
fill level of 1 will generate the same number of elements as the
sparsity pattern of the input matrix while a fill level of 2 will
contain twice as many elements in the factorization, resulting in a
better representation of the true LU factorization. The inverse of the
lower and upper triangular factors may then be easily inverted by
means of simple elimination to produce the preconditioner. For MCSA,
$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
the right to precondition the system.

@@

\subsection{Sparse Approximate Inverse Preconditioning}
\label{subsec:spn_spainv_preconditioning}
SPAINV is chosen because it the true inverse of the preconditioner is
byproduct of the algorithm and the algorithm seeks to maintain the
sparsity of the inverse preconditioner, potentially alleviating some
of the issues observed with ILUT. Explicit formation of the
preconditioner inverse is a requirement of the current MCSA
preconditioning strategy. As a result, although ILUT preconditioning
provided excellent iterative performance, a significant time penalty
is paid to extract the inverse of the triangular factors through $N$
triangular solves. In addition, this explicit inversion yielded a
composite linear operator that was dense, requiring modification of
the system in order to achieve better scalability and Monte Carlo
performance. Sparse approximate inverse (SPAINV) preconditioning is a
technique that may potentially alleviate both of these constraints at
the cost of reduced iterative performance. The method produces a
sparse approximation of the inverse matrix directly by minimizing the
Frobenius norm of the residual matrix \cite{saad_iterative_2003}:
\begin{equation}
  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
  \label{eq:frobenius_norm_min}
\end{equation}
where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
$j^{th}$ columns of the identity and preconditioning matrices
respectively. Here, $\mathbf{M}$ is the actual inverse of the
preconditioner and is formed directly by the algorithm. On the right
hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
represented as a effective sum of $N$ linear system residual norms.

@@
